Solving the independent set problem by generic programming einsum networks \thanks{\funding{...}}

### 2021-08-11 paper/paper.tex
update

%%%%%% Comment related
\newcommand{\red}[1]{[{\bf  \color{red}{LW: #1}}]}
\newcommand{\xred}[1]{[{\bf  \color{red}{\sout{LW: #1}}}]}
\newcommand{\blue}[1]{[{\bf  \color{blue}{JG: #1}}]}
\newcommand{\violet}[1]{[{\bf  \color{violet}{MLS: #1}}]}
\newcommand{\green}[1]{[{\bf  \color{green}{TZ: #1}}]}
\newcommand{\xgreen}[1]{[{\bf  \color{green}{\sout{TZ: #1}}}]}
\newcommand{\xblue}[1]{[{\bf  \color{blue}{\sout{JG: #1}}}]}
\newcommand{\material}[1]{\iffalse[{\bf  \color{cyan}{Material: #1}}]\fi}
\newcommand{\orange}[1]{\iffalse[{\bf  \color{orange}{Jo: #1}}]\fi}

@@

\title{Solving the maximum independant set problem by generic programming einsum networks
\thanks{\funding{...}}
}

@@

\begin{abstract}
	Solving the maximum independent set size problem by mapping the graph to an einsum network. 
    We show how to obtain the maximum independent set size, the independence polynomial and optimal configurations of a graph by engineering the tensor element algebra.
    We also show how to analyse the local properties of a graph by contracting an open einsum network.
\end{abstract}

@@

\section{Introduction}
In this work, we introduce a tensor based framework to study the famous graph problem of finding independent sets.
Given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
Finding the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$ belongs to the complexity class NP-complete~\cite{Hastad1996}, which is unlikely to be decided in polynomial time.
It is hard to even approximate this size in polynomial time within a factor $|V|^{1-\epsilon}$ for an arbituarily small possitive $\epsilon$.
The naive algorithm of enumerating all configuration space gives a $2^{|V|}$ time solution.
More efficient algorithms to compute the MIS size exactly includes the branching algorithm and dynamic programming.
Without changing the fact of exponential scaling in computing time, the branching algorithm gives a smaller base.
For example, in ~\cite{Xiao2017}, a sophisticated branching algorithm gives a time complexity $1.1893^n n ^{O(1)}$.
The dynamic programming approach~\cite{Courcelle1990,Fomin2013} works better for graphs with small tree width $tw(G)$, it gives an algorithms of complexity $O(2^{tw(G)}tw(G)n)$.
People are interested in solving the independent set problem better not only because it is a NP-complete problem that directly related to other NP-complete prolems like maximal cliques and vertex cover~\cite{Moore2011},
but also for its close relation with physical applications like hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.
However, in these applications, knowing the MIS size and one of the optimal solution is not the only goal.
%A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
%A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
People often ask different questions about independent sets in order to understand the landscape of their models better.
These questions includes but not limited to, counting all independent sets, obtaining all indenepent sets of size $\alpha(G)$ and $\alpha(G)-1$,
counting the number of (maximal) independent sets of different sizes, and understanding the effect of a local gadget.
In this work, we attack this problem by mapping it to an generic ``einsum'' network.
It does not give a better time complexity comparing to dynamic programming, but is versatile enough to answer the above questions by engineering the tensor elements with minimum effort.

\section{Einsum network}
%The word ``einsum'' is a shorthand for Einstein's summation, however, modern einsum notation in program is actually invented by a group of programmers.
Einstein's notation is originally proposed as a generalization to of binary matrix multiplication to n-ary tensor contraction.
Let $A, B$ be two matrices, the matrix multiplication is defined as $C_{ik} = \sum_{j}A_{ij}B_{jk}$.
In Einstein's notation, it is denoted as $C_i^k = A_i^j B_j^k$, where the paired subscript and superscript $j$ is a dummy index summed over, hence each index appears precisely twice.
When we have multiple tensors doing the above sum-product operation, we get a tensor network~\cite{Orus2014}.
A tensor network has a nice a muti-graph with open edges. We view a tensor on the right hand side as a vertex in a graph, a label pairing two tensors as an edge, and the remaining unpaired labels as open edges.

\begin{example}
    A tensor networks $C_i^l = A_{ij}^k B^l_k V^j$ has the following graphical representation.

\centerline{\begin{tikzpicture}
    \def\dx{0};
    \def\r{0.25}
    \def\ax{0}
    \def\ay{0}
    \def\bx{1}
    \def\by{1}
    \def\cx{1}
    \def\cy{-1}
    \filldraw[fill=black] (\ax+\dx,\ax) circle [radius=\r];
    \filldraw[fill=black] (\bx+\dx,\by) circle [radius=\r];
    \filldraw[fill=black] (\cx+\dx,\cy) circle [radius=\r];
    %\draw [black,thick] (\ax+\dx,\ay) .. controls (0.5*\ax+0.5*\bx+\dx+0.2,0.5*\ay+0.5*\by-0.2) .. (\bx+\dx,\by);
    %\draw [black,thick] (\ax+\dx,\ay) .. controls (0.5*\ax+0.5*\bx+\dx-0.2,0.5*\ay+0.5*\by+0.2) .. (\bx+\dx,\by);
    \draw [black,thick] (\ax+\dx,\ay) -- (\bx+\dx,\by);
    \draw [black,thick] (\ax+\dx,\ay) -- (\cx+\dx,\cy);
    \draw [black,thick] (\ax+\dx,\ay) -- (\ax+\dx-0.8,\ay);
    \draw [black,thick] (\bx+\dx,\by) -- (\bx+\dx+0.8,\by);
    \node[color=white] at (\ax+\dx,\ax) {A};
    \node[color=white] at (\bx+\dx,\by) {B};
    \node[color=black] at (0.5*\ax+0.5*\bx-0.2+\dx,0.5*\ay+0.5\by+0.2) {k};
    \node[color=black] at (0.5*\ax+0.5*\cx+\dx+0.2,0.5*\ay+0.5\cy+0.2) {j};
    \node[color=black] at (\ax+\dx-0.5,\ay-0.2) {i};
    \node[color=black] at (\bx+\dx+0.5,\by-0.2) {l};
    \node[color=white] at (\cx+\dx,\cy) {V};
\end{tikzpicture}}

@@

Einsum network is a generalization of tensor network by not restricting the number of times a label appears,
hence whether an index is a superscript or a subscript makes no sense now.
It is also called sum-product network or factor graph~\cite{Bishop2006} in some contexts.
The graphical representation of an einsum is a hypergraph, where an edge (label) can be shared by an arbituary number of vertices (tensors).

@@

In the main text, we stick to the einsum notation rather than the tensor network notation.
As a note to those who are more familiar with tensor network representation,
although one can easily translate an einsum network to the equivalent tensor network by adding $\delta$ tensors (a generalization of identity matrix to higher order).
It can sometime increase the contraction complexity of a graph. We have an example demonstrating this in \App{app:tensorbad}.

\section{Independence polynomial}
One can encode the independent set problem on graph $G$ to an einsum network by placing a rank one tensor of size $2$ on vertex $i$
\begin{equation}
    W(x_i)_{s_i} = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right)_{s_i},\label{eq:wtensor}
\end{equation}
and a rank two tensor of size $2 \times 2$ on edge $(i,j)$
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},\label{eq:btensor}
\end{equation}
where a tensor index $s_i$ is a boolean variable that being 1 if vertex $i$ is in the independent set, 0 otherwise.
It corresponds to a hyperedge in the hypergraph.
$x_i$ is a variable.
The contraction of such an einsum network gives
\begin{equation}
    A(G, \{x_1,\ldots,x_n\}) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j}.
\end{equation}
Here, the einsum runs over all vertex configurations $\{s_1,\ldots,s_n\}$ and accumulates the product of tensor elements to the scalar output.
Let $x_i = x$, then the product over vertex tensors gives a factor $x^k$, where $k=\sum_i s_i$ is the vertex set size, 
and the product over edge tensors gives a factor $0$ for configurations not being an independent set.
The contraction of this einsum network gives the independence polynomial~\cite{Ferrin2014, Harvey2017} of $G$
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
The benefit of mapping the independent set problem to the einsum network is one can take the advantage of recently developed techiniques in tensor network based quantum circuit simulations ~\cite{Gray2021,Pan2021},
where people evaluate a tensor network by pairwise contracting tensors in a heuristic order.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{tw(G)})$, where $tw(G)$ is the treewidth of the line graph of a tensor network, 
here it corresponds to the original graph $G$ that we mapped from.~\cite{Markov2008}
The pairwise tensor contraction also makes it possible to utilize fast basic linear algebra subprograms (BLAS) functions for certain tensor element types.

\begin{example}
    Mapping a graph (left) to an einsum network, the resulting einsum network is shown in the right panel.
    A vertex is mapped to a hyperedge in the einsum's graphical notation.
    An edge is mapped to an edge tensor.

@@

        \def\dx{0};
        \def\r{0.25cm}
        \filldraw[fill=black] (\dx,0) circle [radius=\r];
        \filldraw[fill=black] (\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,0) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (2.5+\dx,2.5) circle [radius=\r];
        \draw [black,thick] (\dx,0) -- (\dx,1.5);
        \draw [black,thick] (\dx,0) -- (1.5+\dx,0);
        \draw [black,thick] (\dx,1.5) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (\dx,1.5);
        \draw [black,thick] (2.5+\dx,2.5) -- (1.5+\dx,1.5);
        \node[color=white] at (\dx,0) {a};
        \node[color=white] at (\dx,1.5) {b};
        \node[color=white] at (1.5+\dx,0) {c};
        \node[color=white] at (1.5+\dx,1.5) {d};
        \node[color=white] at (2.5+\dx,2.5) {e};
        \def\dx{5};
        \def\r{0.25cm}
        \foreach \x/\y/\e in {0.75/0/ac, 0/0.75/ab, 1.5/0.75/cd, 0.75/1.5/bd, 0.75/0.75/bc, 2/2/de}
            \node[color=white,fill=black,dot=2*\r] at (\x+\dx,\y) (\e) {\scriptsize B};
        \foreach \x/\y/\v in {0/0/a, 0/1.5/b, 1.5/0/c, 1.5/1.5/d, 2.5/2.5/e}
            \node[color=black] at (\x+\dx,\y) (\v) {$s_\v$};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=white,fill=black,dot=\r] at (\x+\dx,\y) (\v\v) {};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=black] at (\x+\dx+0.5,\y) {\scriptsize $W(x_\v)$};
        \draw [cyan,thick] (a) -- (aa);
        \draw [cyan,thick] (a) -- (ab);
        \draw [cyan,thick] (a) -- (ac);
        \draw [blue,thick] (b) -- (bb);
        \draw [blue,thick] (b) -- (ab);
        \draw [blue,thick] (b) -- (bc);
        \draw [blue,thick] (b) -- (bd);
        \draw [red,thick] (c) -- (cc);
        \draw [red,thick] (c) -- (ac);
        \draw [red,thick] (c) -- (bc);
        \draw [red,thick] (c) -- (cd);
        \draw [green,thick] (d) -- (dd);
        \draw [green,thick] (d) -- (bd);
        \draw [green,thick] (d) -- (de);
        \draw [green,thick] (d) -- (cd);
        \draw [orange,thick] (e) -- (ee);
        \draw [orange,thick] (e) -- (de);
    \end{tikzpicture}}
    The contraction of this network can be done in a pairwise order.
    \begin{align*}
        &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a)_{s_a} W(x_b)_{s_b} W(x_c)_{s_c} W(x_d)_{s_d} W(x_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}.\\
        =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e)_{s_e}\right) W(x_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c)_{s_c}\right)\right)\right.\\
        &\phantom{XXX}\left.\left(B_{s_bs_c}\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a)_{s_a}\right)\right)\right)\\
        =&1 + x_a + x_b + x_c + x_d + x_e + x_ax_d + x_ax_e + x_cx_e + x_bx_e
    \end{align*}
\end{example}

Before contracting the einsum network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relations must hold for arbituary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}

@@

\subsection{The polynomial approach}
A straight forward approach to evaluate the independence polynomial is treating the tensor elements as polynomials, and evaluate the polynomial directly.
Let us create a polynomial type, and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a vector $(a_0, a_1, \ldots, a_k) \in R^k$, e.g. $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),\\
    \mymathbb{1} &= (1).\label{eq:polynomial}
    \end{split}
\end{align}
By contracting the einsum network with polynomial type, the final result is the exact representation of the independence polynomial.
In the program, the multiplication can be evaluated efficiently with the convolution theorem.
The only problem of this method is it suffers from a space overhead that propotional to the maximum independant set size because each polynomial requires a vector of such size to store the factors.
In the following subsections, we managed to solve this problem.

>>>>

%%%%%% Comment related
\newcommand{\red}[1]{[{\bf  \color{red}{ST: #1}}]}
\newcommand{\xred}[1]{[{\bf  \color{red}{\sout{ST: #1}}}]}
\newcommand{\green}[1]{[{\bf  \color{green}{XG: #1}}]}
\newcommand{\xgreen}[1]{[{\bf  \color{green}{\sout{XG: #1}}}]}
\newcommand{\blue}[1]{[{\bf  \color{blue}{JG: #1}}]}
\newcommand{\xblue}[1]{[{\bf  \color{blue}{\sout{JG: #1}}}]}
\newcommand{\material}[1]{\iffalse[{\bf  \color{cyan}{Material: #1}}]\fi}

@@

\title{Solving the independent set problem by generic programming einsum networks
\thanks{\funding{...}}
}

@@

\begin{abstract}
	This paper is about solving the independent set problem by eincoding this problem to an einsum network. 
    We show how to obtain the maximum independent set size, the independence polynomial and optimal configurations of a graph by engineering the tensor element algebra.
    We also show how to analyse the local properties of a graph by contracting an open einsum network.
\end{abstract}

@@

\section{Introduction}
In this work, we introduce a tensor based framework to study the famous graph problem of finding independent sets.
Given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
The problem of finding the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$ belongs to the complexity class NP-complete~\cite{Hastad1996}, which is unlikely to be decided in polynomial time.
It is hard to even approximate this size in polynomial time within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
The exhaustive search for a solution costs time $2^{|V|}$.
More efficient algorithms to compute the MIS size exactly includes the branching algorithm and dynamic programming.
Without changing the fact of exponential scaling in computing time, the branching algorithm gives a smaller base.
For example, in ~\cite{Xiao2017}, a sophisticated branching algorithm has a time complexity $1.1893^n n ^{O(1)}$.
The dynamic programming approach~\cite{Courcelle1990,Fomin2013} works better for graphs with small tree width $tw(G)$, it gives an algorithms of complexity $O(2^{tw(G)}tw(G)n)$.
People are interested in solving the independent set problem better not only because it is an NP-complete problem that directly related to other NP-complete problems like maximal cliques and vertex cover~\cite{Moore2011},
but also for its close relation with physical applications like hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.
However, in these applications, knowing the MIS size and one of the optimum solutions is not the only goal.
%A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
%A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
People often ask different questions about independent sets in order to understand the landscape of their models better.
These questions includes but not limited to, counting all independent sets, obtaining all independent sets of size $\alpha(G)$ and $\alpha(G)-1$,
counting the number of (maximal) independent sets of different sizes, and understanding the effect of a local gadget.
In this work, we attack this problem by mapping it to an generic ``einsum'' network.
It does not give a better time complexity comparing to dynamic programming, but is versatile enough to answer the above questions by engineering the tensor elements with minimum effort.

\section{Einsum networks}
%The word ``einsum'' is a shorthand for Einstein's summation, however, modern einsum notation in program is actually invented by a group of programmers.
Einstein's notation is originally proposed as a generalization to of binary matrix multiplication to n-ary tensor contraction.
Let $A, B$ be two matrices, the matrix multiplication is defined as $C_{ik} = \sum_{j}A_{ij}B_{jk}$.
In Einstein's notation, it is denoted as $C_i^k = A_i^j B_j^k$, where the paired subscript and superscript $j$ is a dummy index summed over, hence each index appears precisely twice.
When we have multiple tensors doing the above sum-product operation, we get a tensor network~\cite{Orus2014}.
A tensor network can be represented as a mutigraph with open edges by viewing a tensor as a vertex, a label pairing two tensors as an edge, and the remaining unpaired labels as open edges.

\begin{example}
    A tensor networks $C_i^l = A_{ij}^{km} B^l_{km} V^j$ has the following multigraph representation.

\centerline{\begin{tikzpicture}[
    dot/.style = {circle, fill, minimum size=#1,
                inner sep=0pt, outer sep=0pt},
    dot/.default = 6pt  % size of the circle diameter 
                    ]  
    \def\dx{0};
    \def\r{0.25cm}
    \def\ax{0}
    \def\ay{0}
    \def\bx{1}
    \def\by{1}
    \def\cx{1}
    \def\cy{-1}
    \node[color=white, fill=black, dot=2*\r] at (\ax+\dx,\ax) (a) {A};
    \node[color=white, fill=black, dot=2*\r] at (\bx+\dx,\by) (b) {B};
    \node[color=white, fill=black, dot=2*\r] at (\cx+\dx,\cy) (c) {V};
    %\draw [black,thick] (\ax+\dx,\ay) .. controls (0.5*\ax+0.5*\bx+\dx+0.2,0.5*\ay+0.5*\by-0.2) .. (\bx+\dx,\by);
    %\draw [black,thick] (\ax+\dx,\ay) .. controls (0.5*\ax+0.5*\bx+\dx-0.2,0.5*\ay+0.5*\by+0.2) .. (\bx+\dx,\by);
    \draw [black,thick] (a) .. controls (0.4, 0.6) .. (b);
    \draw [black,thick] (a) .. controls (0.6, 0.4) .. (b);
    \draw [black,thick] (a) -- (c);
    \draw [black,thick] (a) -- (\ax+\dx-0.8,\ay);
    \draw [black,thick] (b) -- (\bx+\dx+0.8,\by);
    \node[color=black] at (0.5*\ax+0.5*\bx-0.3+\dx,0.5*\ay+0.5\by+0.3) {k};
    \node[color=black] at (0.5*\ax+0.5*\bx+0.3+\dx,0.5*\ay+0.5\by-0.3) {m};
    \node[color=black] at (0.5*\ax+0.5*\cx+\dx+0.2,0.5*\ay+0.5\cy+0.2) {j};
    \node[color=black] at (\ax+\dx-0.5,\ay-0.2) {i};
    \node[color=black] at (\bx+\dx+0.5,\by-0.2) {l};
\end{tikzpicture}}

@@

Einsum network is a generalization of tensor network by not restricting the number of times a label appears in the notation,
hence whether an index is a superscript or a subscript makes no sense now.
It is also called sum-product network or factor graph~\cite{Bishop2006} in some contexts.
The graphical representation of an einsum is a hypergraph, where an edge (label) can be shared by an arbitrary number of vertices (tensors).

@@

In the main text, we stick to the einsum notation rather than the tensor network notation.
As a note to those who are more familiar with tensor network representation,
although one can easily translate an einsum network to the equivalent tensor network by adding $\delta$ tensors (a generalization of identity matrix to higher order).
It can sometime increase the contraction complexity of a graph. We have an example demonstrating this point in \App{app:tensorbad}.

\section{Independence polynomial}
One can encode the independence polynomial~\cite{Ferrin2014, Harvey2017} of $G$ to an einsum network.
Independence polynomial is an important graph polynomial that contains the counting information of an independent set problem. It is defined as
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
We encode this polynomial to an einsum network by placing a rank one tensor of size $2$ parametrized by $x_i$ on a vertex $i$
\begin{equation}
    W(x_i)_{s_i} = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right)_{s_i},\label{eq:wtensor}
\end{equation}
and a rank two tensor of size $2 \times 2$ on an edge $(i,j)$
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},\label{eq:btensor}
\end{equation}
where a tensor index $s_i$ is a boolean variable that having the meaning of being 1 if vertex $i$ is in the independent set, 0 otherwise.
It corresponds to a hyperedge in the hypergraph.
The contraction of such an einsum network gives
\begin{equation}
    P(G, \{x_1,\ldots,x_n\}) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1,\ldots,s_n\}$ and accumulates the product of tensor elements to the scalar output $P$.
We can see an edge tensor represents the restriction on an edge that if both vertices connected by it are included in the set, then such configuration has no contribution to the output.
When we set $x_i = x$, the contraction result corresponds to the independence polynomial.
One can see the connection from the fact that the product over vertex tensor elements gives a factor $x^k$, where $k=\sum_i s_i$ counts the set size, 
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set, $0$ otherwise.
One directly benefit of mapping the independent set problem to an einsum network is one can take the advantage of recently developed techniques in tensor network based quantum circuit simulations ~\cite{Gray2021,Pan2021},
where people evaluate an einsum network by pairwise contracting tensors in a heuristic order.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{tw(G)})$.
Here $tw(G)$ is the tree width of the line graph of an einsum hypergraph, while the line graph of an einsum hypergraph corresponds to the original graph $G$ that we mapped from.~\cite{Markov2008}
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up our computation for certain tensor element types.

\begin{example}
    Mapping a graph (left) to an einsum network, the resulting einsum network is shown in the right panel.
    In the einsum's graphical representation, a vertex is mapped to a hyperedge,
    and an edge is mapped to an edge tensor.

@@

        \def\dx{0};
        \def\r{0.25cm}
        \filldraw[fill=black] (\dx,0) circle [radius=\r];
        \filldraw[fill=black] (\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,0) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (2.5+\dx,2.5) circle [radius=\r];
        \draw [black,thick] (\dx,0) -- (\dx,1.5);
        \draw [black,thick] (\dx,0) -- (1.5+\dx,0);
        \draw [black,thick] (\dx,1.5) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (\dx,1.5);
        \draw [black,thick] (2.5+\dx,2.5) -- (1.5+\dx,1.5);
        \node[color=white] at (\dx,0) {a};
        \node[color=white] at (\dx,1.5) {b};
        \node[color=white] at (1.5+\dx,0) {c};
        \node[color=white] at (1.5+\dx,1.5) {d};
        \node[color=white] at (2.5+\dx,2.5) {e};
        \def\dx{5};
        \def\r{0.25cm}
        \foreach \x/\y/\e in {0.75/0/ac, 0/0.75/ab, 1.5/0.75/cd, 0.75/1.5/bd, 0.75/0.75/bc, 2/2/de}
            \node[color=white,fill=black,dot=2*\r] at (\x+\dx,\y) (\e) {\scriptsize B};
        \foreach \x/\y/\v in {0/0/a, 0/1.5/b, 1.5/0/c, 1.5/1.5/d, 2.5/2.5/e}
            \node[color=black] at (\x+\dx,\y) (\v) {$s_\v$};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=white,fill=black,dot=\r] at (\x+\dx,\y) (\v\v) {};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=black] at (\x+\dx+0.5,\y) {\scriptsize $W(x_\v)$};
        \draw [cyan,thick] (a) -- (aa);
        \draw [cyan,thick] (a) -- (ab);
        \draw [cyan,thick] (a) -- (ac);
        \draw [blue,thick] (b) -- (bb);
        \draw [blue,thick] (b) -- (ab);
        \draw [blue,thick] (b) -- (bc);
        \draw [blue,thick] (b) -- (bd);
        \draw [red,thick] (c) -- (cc);
        \draw [red,thick] (c) -- (ac);
        \draw [red,thick] (c) -- (bc);
        \draw [red,thick] (c) -- (cd);
        \draw [green,thick] (d) -- (dd);
        \draw [green,thick] (d) -- (bd);
        \draw [green,thick] (d) -- (de);
        \draw [green,thick] (d) -- (cd);
        \draw [orange,thick] (e) -- (ee);
        \draw [orange,thick] (e) -- (de);
    \end{tikzpicture}}
    The contraction of this network can be done in a pairwise order.
    \begin{align*}
        &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a)_{s_a} W(x_b)_{s_b} W(x_c)_{s_c} W(x_d)_{s_d} W(x_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}.\\
        =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e)_{s_e}\right) W(x_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c)_{s_c}\right)\right)\right.\\
        &\phantom{XXX}\left.\left(B_{s_bs_c}\left(\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a)_{s_a}\right)\right)\right)\right)\\
        =&1 + x_a + x_b + x_c + x_d + x_e + x_ax_d + x_ax_e + x_cx_e + x_bx_e\\
        =&1+5x+4x^2
    \end{align*}
\end{example}

Before contracting the einsum network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relations must hold for arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}

@@

\subsection{The polynomial approach}
A straight forward approach to evaluate the independence polynomial is treating the tensor elements as polynomials, and evaluate the polynomial directly.
Let us create a polynomial type, and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a vector $(a_0, a_1, \ldots, a_k) \in R^k$, e.g. $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),\\
    \mymathbb{1} &= (1).\label{eq:polynomial}
    \end{split}
\end{align}
By contracting the einsum network with polynomial type, the final result is the exact representation of the independence polynomial.
In the program, the multiplication can be evaluated efficiently with the convolution theorem~\cite{Schonhage1971}.
The only problem of this method is it suffers from a space overhead that proportional to the maximum independent set size because each polynomial requires a vector of such size to store the factors.
In the following subsections, we managed to solve this problem.

\subsection{The fitting and Fourier transformation approaches}
Let $m=\alpha(G)$ be the maximum independent set size and $X$ be a set of real numbers of cardinality $m+1$.
We compute the einsum contraction for each $x_i \in X$ and obtain the following relations
\begin{align}
    \begin{split}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
    \end{split}
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.
The polynomial fitting is essentially about solving the following linear equation
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).\label{eq:lineareq}
\end{align}
